```{r setup}
install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will be practicing with dplyr functions using the starwars dataset:

```{r}
starwars
head(starwars)
```

# Subset rows with filter

filter() allows

```{r}
##Filter for rows with human Star Wars for rows with human Star Wars chracters
filter(starwars, species == "Human")
humans
filter(humans, sex == "male")
humans <- filter(starwars, species == "Human", sex == "male")
filter(starwars, species == "Human", sex == "male")
```

We can use different operators to filter:

```{r}
##return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all chracters that are NOT human?
filter(starwars, species != "Human")
```

variable_name %in% vector_of_values returns TRUE if variable_name is equal to one of the values in vector_of_values

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan" )

filter(starwars, hair_color %in% c("brown", "blond"))

```

grepel("characters_of_intrest", variable_name) returns TRUE if variable_name contains the characters_of_intrest

```{r}
filter(starwars, grepl("blond", hair_color))

filter(starwars, grepl("blue", eye_color))
```

You can use \<, \>, \<=, \>= to comapre numerica variables or categorical variables

-   nominal categorical variables are ranked alphabetically,

-   ordinal categorical variables (factors are ranked based on the built-in order

```{r}
filter(starwars, name <"D") ## ranking based on alphabetical order

str(diamonds)
filter(diamonds, cut > "Good") #ranking is based on leveels of ordered factor cut

"D" < "d" #false
"d" < "D" #true
```

between(x, left, right) is a shorthand for x \>= left & x \<=right

```{r}
between(5, 1, 7)
filter(starwars, between(mass, 50, 100))
```

## Practice

```{r}
filter(starwars, is.na(hair_color))

filter(starwars, height < 100, species != "Droid")

filter(starwars, between(birth_year, 100, 200))

```

# Order rows with arrange

arrange() reorders rows based on the values of column(s)

```{r}
##order by name
arrange(starwars, name)

#order by species and birth year
arrange(starwars, species, birth_year)

#birth year in descending
arrange(starwars, species, desc(birth_year))
```

```         
```

Practice

```{r}

arrange(starwars, desc(height))

arrange(starwars, height, species) 
arrange(starwars, species, height) 

arrange(starwars, hair_color, eye_color)



```

# Subset columns with select()

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
#all columns between name and eye_color (inclusive)
select(starwars, name:eye_color)
```

We can also remove columns by name using a -

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
```

Helpful functions

everything0 is a shortcut for "everything else"

```{r}
select(starwars, species, everything())
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)

```

```{r}
select(starwars, starts_with("h"))

select(starwars, ("or"))
```

Practice

```{r}
select(starwars, homeworld, species)

select(starwars, contains("color"), birth_year)
select(starwars, hair_color:birth_year)
select(starwars, hair_color, skin_color, eye_color )
```

# Add new variables with mutate

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: sum(), mean()

```{r}
mutate(starwars_small,
      row_addtion = height + mass,
      summary_sum = sum(height, mass, na.rm = T))
```

Group by a variable using the.by parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na,rm=T),
       .by = species)
```

Practice

```{r}
mutate(starwars_small,
       mass_lbs = mass*2.2)
```

Summarize data with summarize

```{r}
summarize(starwars,
         mean_mass = mean(mass, na.rm = T))

summarize(starwars,
         mean_mass = mean(mass, na.rm = t),
         .by = species)
```
